---
title: "Deep learning, animal species identification, camera traps, spatial predator-prey co‐occurrence"
author: "Olivier Gimenez, Maëlis Kervellec, Anna Chaine, Jean-Baptiste Fanjul, Christophe Duchamp (who else?)"
date: "`r Sys.Date()`"
always_allow_html: true
output:
  html_document:
    highlight: tango
    theme: yeti
    toc: yes
    toc_depth: 2
    number_sections: yes
link-citations: yes
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE, fig.width = 8, fig.height = 8, echo = FALSE)
options(htmltools.dir.version = FALSE)
```

<button type="button" class="btn" title="Print to PDF" onClick="window.print()">Export to PDF</button>

# Abstract

Blabla. 

# Introduction

+ Explosion of computer vision applications. 

+ In ecology in particular to do this and that. Make sure to cite latest review papers. 

+ However, two communities bit disconnected despite efforts by people at the interface, and platforms like Wildlife Insights. 

+ Might explain why papers seldom propose the entire workflow of deep learning and ecological question treatment. 

+ Also difficult to apply for non specialists, different terminology but also Python vs. R. 

+ And lack of reproducibility. 

+ Here, show case complete workflow on a case study on predator prey interactions. Data retrieving (pix), deep learning, then quantification of spatial co-occurrence. We argue that maybe not worth spending valuable time tuning algo to increase perf by one or two percent, if ecological indicators not too much affected. Entirely in R with which ecologists are familiar, and entirely reproducible. 

# Motivating case study

+ Lynx and its prey

+ PPP Lynx

+ Data collection

+ Main objective

# Methods 

## Data curation

## Species identification

### Training with Jura

### Validation with Ain

## Spatial co-occurrence

# Results

# Discussion

+ Summarise what we did. 

+ Main lessons.

+ To be extended. 

## Bibliographic references

## Acknowledgments

ANR. 

# Session information

```{r session-info}
sessionInfo()
```

# References
