---
title: "Trade-off between deep learning for species identification and inference about predator-prey co-occurrence: Reproducible `R` workflow integrating models in computer vision and ecological statistics"
author: "Olivier Gimenez, Maëlis Kervellec, Jean-Baptiste Fanjul, Anna Chaine, Lucile Marescot, Yoann Bollet, Christophe Duchamp"
date: "`r Sys.Date()`"
always_allow_html: true
output:
  html_document:
    highlight: tango
    theme: yeti
    toc: yes
    toc_depth: 2
    number_sections: yes
link-citations: yes
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE, 
                      fig.width = 8, 
                      fig.height = 8, 
                      echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      cache = FALSE)
options(htmltools.dir.version = FALSE)
set.seed(2022) # because 2019, 2020 and 2021 were not great
library(tidyverse)
theme_set(theme_light(base_size = 14))
library(fastai)
library(kableExtra)
```

## Transfer learning

We proceed with transfer learning using images from the Jura county (or a subsample more exactly). We first load images and apply standard transformations to improve training (flip, rotate, zoom, rotate, ligth transform).
```{r echo = TRUE}
dls <- ImageDataLoaders_from_folder(
  path = "pix/",
  train = "train",
  valid = "valid",
  item_tfms = Resize(size = 460),
  bs = 10,
  batch_tfms = list(aug_transforms(size = 224,
                                   min_scale = 0.75), # transformation
                    Normalize_from_stats( imagenet_stats() )),
  num_workers = 0,
  ImageFile.LOAD_TRUNCATED_IMAGES = TRUE)
```

Then we get the model architecture. For the sake of illustration, we use a resnet18 here, but we used a resnet50 to get the full results presented in the main text.
```{r echo = TRUE}
learn <- cnn_learner(dls = dls,
                     arch = resnet18(),
                     metrics = list(accuracy, error_rate))
```

Now we are ready to train our model. Again, for the sake of illustration, we use only 2 epochs here, but used 20 epochs to get the full results presented in the main text. With all pictures and a resnet50, it took 75 minutes per epoch approximatively on a Mac with a 2.4Ghz processor and 64Go memory, and less than half an hour on a machine with GPU. On this reduced dataset, it took a bit more than a minute per epoch on the same Mac. Note that we save the model after each epoch for later use.
```{r, echo = TRUE, cache = FALSE}
one_cycle <- learn %>%
  fit_one_cycle(2, cbs = SaveModelCallback(every_epoch = TRUE,
                                           fname = 'model'))
one_cycle
```

We may dig a bit deeper in training performances by loading the best model, here `model_1.pth`, and display some metrics for each species.
```{r echo = TRUE}
learn$load("model_1")
interp <- ClassificationInterpretation_from_learner(learn)
interp$print_classification_report()
```

We may extract the categories that get the most confused.
```{r echo = TRUE}
interp %>% most_confused()
```

## Transferability

In this section, we show how to use our freshly trained model to label images that were taken in another study site in the Ain county, and not used to train our model. First, we get the path to the images.
```{r echo = TRUE}
fls <- list.files(path = "pix/pixAin",
                  full.names = TRUE,
                  recursive = TRUE)
```

Then we carry out prediction, and compare to the truth.
```{r echo = TRUE}
predicted <- character(3)
categories <- interp$vocab %>%
  str_replace_all("[[:punct:]]", " ") %>%
  str_trim() %>%
  str_split("   ") %>%
  unlist()
for (i in 1:length(fls)){
  result <- learn %>% predict(fls[i]) # make prediction
  result[[3]] %>%
    str_extract("\\d+") %>%
    as.integer() -> index # extract relevant info
  predicted[i] <- categories[index + 1] # match it with categories
}
data.frame(truth = c("lynx", "roe deer", "wild boar"),
           prediction = predicted) %>%
  kable(caption = "Comparison of the predictions vs. ground truth.") %>%
  kable_styling()
```
